WasteAndMaterialFootprint: A python package to quantify supply chain flows of waste and material in LCA databases

### 2024-01-31 latex_files/discussion.tex
update-manuscript

\begin{itemize}
    \item Summary of the results, 1 par
    \item -----------------------------
    \item Advantages of the WMF tool, 1 par
    \item -----------------------------
    \item Limitations, 1 par
    \item -----------------------------
    \item Future possibilities, 1 par
\end{itemize}

>>>>

Given that both waste generation and material demand are often strongly associated with the environmental impacts of an activity, it is important that they are included in the LCA. While there are numerous examples of existing and proposed methods that attempt to provide endpoint LCIA scores through convoluted formulae or subjective weighting, there is little consensus on their application and their complexity and lack of transparency can make them difficult to use and interpret.

The WMF tool, in contrast, gives LCA practitioners a simple, flexible, and transparent way to calculate the supply chain waste and material footprints, delivering results in standard units as direct aggregations of the relevant demand inventories. Once the databases have been processed with the WMF tool, the user can then easily apply the WMF methods to calculate the supply-chain waste and material footprints for any activity in the same way that they would calculate any LCIA indicator.

The simple case study presented in this paper demonstrates both the utility and the limitations of the WMF tool. It was shown the WMF tool was able to calculate categorised aggregates and contribution analyses of both the waste generated and material demands in the present and future supply chains of the Li-ion batteries under consideration. Further, integration of the WMF methods as `pseudo-LCIA' methods allows the user to easily make use of the WMF tool in their preferred LCA software, be it code-based like \texttt{brightway2} or graphical like \texttt{ActivityBrowser}.

One main limitation of the WMF tool is that it does not yet provide specific information (in a readily accessible format) on the composition of the waste generated, which would be needed to thoroughly assess the environmental impacts of this waste. Currently, the user would need to manually explore the `waste inventory' produced by the application of the WMF to determine if, for example, the waste generated represents an actual loss of resources, or is simply a transfer of the `overburden' in mining activity, which is classified as `inert waste'. A methodic classification of waste exchanges and the end-of-life fates will be facilitated by the more detailed and disaggregated data that is seen in each successive release of ecoinvent~\citep{fitzgerald2023ecoinventdocumentation}.

Furthermore, the utility of the WMF tool in studies of future supply chains is limited by the fact that the currently available prospective databases focus largely on changes in the energy, steel, cement, and transport sectors~\citep{sacchi2023premisedocs}. As demonstrated in the results of the case study---where there was often very little scenario-temporal change in many waste and material footprint indicators---the utility of prospective LCA is restricted if there is little adaption of the future background inventories. In particular, the inclusion of scenarios with future waste processing technology would greatly improve our predictions of waste and material flows. A strong focus on enhancing these prospective databases is, thus, of critical importance to the future of prospective LCA, and by extension, to the development of the circular economy.

### 2024-01-31 latex_files/main.tex
update-manuscript

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
\documentclass[review,3p,authoryear]{elsarticle}
%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}
\usepackage{lineno}
\usepackage{booktabs}
% ---------------------------------------------------------------------
%             FOR DRAFTING / REVIEW (DELETE BEFORE SUBMISSION)
% ---------------------------------------------------------------------
\usepackage{caption}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage[tt=false]{libertine}
\usepackage{xcolor}
\definecolor{blue}{RGB}{33,150,209}
\definecolor{red}{RGB}{255,0,0}
\usepackage{xurl}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{pdfpages}
\usepackage{enumitem}
\usepackage{lineno}
\usepackage{float}
\setlength{\linenumbersep}{5pt}
% Save the old \texttt command
\let\oldtexttt\texttt
% Redefine the \UrlFont to use a smaller typewriter font
\renewcommand{\UrlFont}{\ttfamily\small}
% Redefine the \texttt command to use a smaller font size and allow line breaks
\renewcommand{\texttt}[1]{{\ttfamily\small\nolinkurl{#1}}}

@@

\section{Discussion}
\label{sec:discussion}
\input{discussion.tex}
\cbox{Section word count: aim for 400?}

\section{Conclusions}
\label{sec:conclusions}
\cbox{Section word count: 250}
We have written the WasteAndMaterialFootprint (WMF) tool, an extension to the brightway2 LCA framework that enables the user to calculate the waste and material footprints of a product or service in an LCA database. It explodes the database, identifies upstream waste and material exchanges, edits them and writes matching custom WMF methods. These exchanges become pseudo-biosphere flows and thus, the footprint can be calculated as with the existing LCIA methods. The WMF tool can be easily customised by the user to calculate the footprints of other supply-chain flows such as water, gas, and critical raw materials.

Given that both waste generation and material demand are often strongly associated with the environmental impacts of an activity, it is important that they are included in the LCA. While there are numerous examples of existing and proposed methods that attempt to provide endpoint LCIA scores through convoluted formulae or subjective weighting, there is little consensus on their application and their complexity and lack of transparency can make them difficult to use and interpret.

The WMF tool, in contrast, gives LCA practitioners a simple, flexible, and transparent way to calculate the supply chain waste and material footprints, delivering results in standard units as direct aggregations of the relevant demand inventories.

@@

\section*{Supplementary material}
    The supplementary material contains the following:
    \begin{enumerate}
        \item List of waste and material categories
        \item Example output of the WasteAndMaterialFootprint tool
        \item Extended results of the case study
    \end{enumerate}

>>>>

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
\documentclass[final,5p,authoryear]{elsarticle}
%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}
\usepackage{lineno}
\usepackage{booktabs}
% ---------------------------------------------------------------------
%             FOR DRAFTING / REVIEW (DELETE BEFORE SUBMISSION)
% ---------------------------------------------------------------------
\usepackage{caption}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage[tt=false]{libertine}
\usepackage{xcolor}
\definecolor{blue}{RGB}{33,150,209}
\definecolor{red}{RGB}{255,0,0}
\usepackage{xurl}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{pdfpages}
\usepackage{enumitem}
\usepackage{lineno}
\usepackage{float}
\setlength{\linenumbersep}{5pt}
% Save the old \texttt command
\let\oldtexttt\texttt
% Redefine the \UrlFont to use a smaller typewriter font
\renewcommand{\UrlFont}{\ttfamily\small}
% Redefine the \texttt command to use a smaller font size and allow line breaks
\renewcommand{\texttt}[1]{{\ttfamily\small\nolinkurl{#1}}}

@@

\section{Discussion}
\label{sec:discussion}
\input{discussion.tex}
\cbox{Section word count: 450}

\section{Conclusions}
\label{sec:conclusions}
\cbox{Section word count: 250}
We have written the WasteAndMaterialFootprint (WMF) tool, an extension to the brightway2 LCA framework that enables the user to calculate the waste and material footprints of a product or service in an LCA database. It explodes the database, identifies upstream waste and material exchanges, edits them, and writes matching custom WMF methods. These exchanges become pseudo-biosphere flows and thus, the footprint can be calculated as with the existing LCIA methods. The WMF tool can be easily customised by the user to calculate the footprints of other supply-chain flows such as water, gas, and critical raw materials.

@@

\section*{Supplementary material}
    The supplementary material contains the following:
    \begin{enumerate}
        \item List of waste and material categories
        \item Example output of the WasteAndMaterialFootprint tool
        \item List of identified waste and material exchanges
        \item Code (python script) used for case study
        \item Inventory and methods used in the case study
        \item Complete tabulated results of the case study
        \item Complete visualisations of the case study
    \end{enumerate}

### 2024-01-31 latex_files/methodology.tex
update-manuscript

This section is divided into two parts. In \autoref{sec:method-WMF_tool}, we describe the WasteAndMaterialFootprint tool, in \autoref{sec:method-case_study}, we describe the methodology used to calculate the waste and material footprints in the case study.

\subsection{The WasteAndMaterialFootprint tool}
\label{sec:method-WMF_tool}

The WasteAndMaterialFootprint tool is a Python package that allows one to calculate the waste and material footprint of any product or service inside of LCA databases. The tool is built on the \texttt{Brightway2} LCA framework~\citep{mutel2017brightway} and is also compatible with \texttt{ActivityBrowser}~\citep{steubing2020activitybrowser} an open-source graphical user interface for LCA. The WMF tool is installable via the Python Package Index (PyPI)~\citep{mcdowall2023wmfpipy} and is open source under the CC-0 licence. The full source code for the WasteAndMaterialFootprint tool is indexed on Zenodo~\citep{mcdowall2023wmfzenodo} and under further development in the GitHub repository~\citep{mcdowall2024wmfgithub}. The tool is designed to be used with ecoinvent databases~\citep{ecoinvent2016version3}, but could be adapted to other databases as well by changing the search criteria. Currently, it has been tested with all available system models of ecoinvent 3.5--3.10.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\columnwidth]{figures/WMF_flowchart.pdf}
    \caption{Flowchart of the WasteAndMaterialFootprint tool \cbox{Figure needs to be updated with new names + premise}}\label{fig:WMF_flowchart}
\end{figure}

>>>>

This section is divided into two parts. In \autoref{sec:method-WMF_tool}, we describe the WasteAndMaterialFootprint (WMF) tool, in \autoref{sec:method-case_study}, we describe the methodology used to calculate the waste and material footprints in the case study.

\subsection{The WasteAndMaterialFootprint (WMF) tool}
\label{sec:method-WMF_tool}

The WMF tool is a Python package that allows one to calculate the waste and material footprint of any product or service inside of LCA databases. Metadata of the WMF tool is presented in \autoref{tab:methods-wmfmetadata}.

\begin{table}[h]
    \caption{WasteAndMaterialFootprint (WMF) tool metadata}
    \label{tab:methods-wmfmetadata}
    \centering
    \begin{tabular}{ll}
    \toprule
    \textbf{Item} & \textbf{Details} \\
    \midrule
    Current version & 0.1.21 \\
    DOI & \url{zenodo.org/doi/10.5281/zenodo.10431180} \\
    Code repository & \url{github.com/Stew-McD/WasteAndMaterialFootprint} \\
    License & CC0--1.0 license \\
    Versioning system & git \\
    Language & Python \\
    Documentation & \url{wasteandmaterialfootprint.readthedocs.io} \\
    Main dependencies & brightway2, premise, wurst \\
    Support email & s.c.mcdowall@cml.leidenuniv.nl \\
    \bottomrule
    \end{tabular}
\end{table}

The tool is built on the \texttt{Brightway2} LCA framework~\citep{mutel2017brightway} and is also compatible with \texttt{ActivityBrowser}~\citep{steubing2020activitybrowser} an open-source graphical user interface for LCA. The WMF tool is installable via the Python Package Index (PyPI)~\citep{mcdowall2023wmfpipy} and is open source under the CC-0 licence. The full source code for the WMF tool is indexed on Zenodo~\citep{mcdowall2023wmfzenodo} and under further development in the GitHub repository~\citep{mcdowall2024wmfgithub}. The tool is designed to be used with ecoinvent databases~\citep{ecoinvent2016version3}, but could be adapted to other databases as well by changing the search criteria. Currently, it has been tested with all available system models of ecoinvent 3.5--3.10.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[height=0.8\textheight]{figures/WMF_flowchart.pdf}
    \caption{Flowchart of the WasteAndMaterialFootprint (WMF) tool}\label{fig:WMF_flowchart}
\end{figure}

### 2024-01-31 latex_files/results.tex
update-manuscript

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{figures/top-processes.png}
    \caption{Contribution of `top-processes' to the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:top_contribution}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\linewidth]{figures/comparison_methods.png
    }
    \caption{Comparison of LCIA method `EDIP 2003 - coal no LT' with the `WMF - Coal (black)' in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model. }\label{fig:comparison_methods}
\end{figure}

>>>>

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{figures/top-processes.png}
    \caption{Contribution of `top-processes' to the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:top_contribution}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.8\linewidth}
        \includegraphics[width=\linewidth]{figures/comparison_methods2.png}
        \caption{}\label{fig:comparison_methods_a}
    \end{subfigure}

    \begin{subfigure}[b]{0.8\linewidth}
        \includegraphics[width=\linewidth]{figures/comparison_methods1.png}
        \caption{}\label{fig:comparison_methods_b}
    \end{subfigure}
    \caption{Comparison of LCIA method `EDIP 2003 - coal no LT' (a) with the `WMF - Coal (black)' (b) in the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:comparison_methods}
\end{figure}
